---
title: eDNA analysis Everglades
output: md_document
editor_options: 
  chunk_output_type: console
---

# eDNA analysis Everglades

This notebook contains an exploratory data analysis for Everglades National Park. Images and tables are exported to the `report`` folder.

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=300)
library(ggplot2)
library(dplyr)
library(purrr)
library(stringr)
library(xtable)
library(rredlist)
library(sf)
library(ggsvg)
library(ggrepel)
library(worrms)
```

## Read dataset

```{r message=FALSE, warning=FALSE}
dna_files <- list.files("edna-results/data", "*DNADerivedData*", full.names = TRUE)
dna_files <- dna_files[str_detect(dna_files, "everglades")]
occurrence_files <- list.files("edna-results/data", "*Occurrence*", full.names = TRUE)
occurrence_files <- occurrence_files[str_detect(occurrence_files, "everglades")]

dna <- map(dna_files, read.table, sep = "\t", quote = "", header = TRUE) %>%
  bind_rows() %>%
  mutate_if(is.character, na_if, "")

occurrence <- map(occurrence_files, read.table, sep = "\t", quote = "", header = TRUE) %>%
  bind_rows() %>%
  mutate_if(is.character, na_if, "") %>%
  mutate(species = ifelse(taxonRank == "species", scientificName, NA)) %>%
  mutate(AphiaID = as.numeric(str_replace(scientificNameID, "urn:lsid:marinespecies.org:taxname:", ""))) %>%
  left_join(dna, by = "occurrenceID") %>%
  filter(higherGeography == "Everglades National Park")
```

## Cleanup

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
cleanup <- read.csv("species_lists/cleanup.csv") %>%
  mutate(remove = as.logical(remove))

aphiaid_remove <- cleanup %>% filter(remove) %>% pull(AphiaID)

occurrence <- occurrence %>%
  filter(!AphiaID %in% aphiaid_remove)

aphiaid_replace <- cleanup %>%
  filter(!is.na(new_AphiaID))

taxa_replace <- wm_record(aphiaid_replace$new_AphiaID) %>%
  select(AphiaID, phylum, class, order, family, genus, species = scientificname, scientificName = scientificname, scientificNameID = lsid, taxonRank = rank) %>%
  mutate(taxonRank = tolower(taxonRank)) %>%
  mutate(old_AphiaID = aphiaid_replace$AphiaID)

occurrence <- occurrence %>%
  mutate(old_AphiaID = AphiaID) %>%
  rows_update(taxa_replace, by = "old_AphiaID") %>%
  select(-old_AphiaID)
```

## Overall statistics

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
sample_volumes <- occurrence %>%
  distinct(materialSampleID, sampleSize) %>%
  mutate(sampleSize = as.integer(sampleSize))
  
sample_stats <- occurrence %>%
  group_by(locality, materialSampleID) %>%
  summarize(reads = sum(organismQuantity), species = n_distinct(na.omit(species)), asvs = n_distinct(DNA_sequence)) %>%
  left_join(sample_volumes, by = "materialSampleID")

stats <- occurrence %>%
  summarize(reads = sum(organismQuantity), species = n_distinct(na.omit(species)), asvs = n_distinct(DNA_sequence))

marker_stats <- occurrence %>%
  group_by(pcr_primer_name_forward) %>%
  summarize(reads = sum(organismQuantity), species = n_distinct(na.omit(species)), asvs = n())

sample_stats_x <- xtable(sample_stats, type = "latex")
caption(sample_stats_x) <- "Reads, ASVs, and species by sample."
label(sample_stats_x) <- "table:samples"
print(sample_stats_x, file = "report/sample_stats.tex", include.rownames = FALSE, tabular.environment = "longtable", floating = FALSE)

stats_x <- xtable(stats, type = "latex")
caption(stats_x) <- "Reads, ASVs, and species across all samples."
label(stats_x) <- "table:stats"
print(stats_x, file = "report/stats.tex", include.rownames = FALSE, tabular.environment = "longtable", floating = FALSE)
```

## Sampling location map

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
everglades <- read_sf("map/everglades.gpkg")

sites <- occurrence %>%
  select(decimalLongitude, decimalLatitude, locality) %>%
  distinct() %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326, remove = FALSE)

ggplot() +
  geom_sf(data = everglades %>% st_transform(4326), fill = NA, color = alpha("#237d2a", 0.5), lwd = 2) +
  geom_sf(data = st_as_sfc("POLYGON ((-82 26, -80 26, -80 24.8, -82 24.8, -82 26))", crs = 4326), fill = NA, color = NA) +
  geom_point(data = sites, aes(decimalLongitude, decimalLatitude), color = "#e3007d", shape = 21, stroke = 2, size = 3) +
  geom_text_repel(data = sites, aes(decimalLongitude, decimalLatitude, label = locality), size = 5, box.padding = 2, point.padding = 1, force_pull = 0.1) +
  landr::geom_landr(after = 0, fill = "#f2f2f2") +
  theme_void()

ggsave("report/map.png", scale = 1.2, bg = "white", width = 10, height = 7)
```

## Species list

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
species <- occurrence %>%
  filter(!is.na(species)) %>%
  mutate(AphiaID = as.numeric(str_replace(scientificNameID, "urn:lsid:marinespecies.org:taxname:", ""))) %>%
  group_by(phylum, class, order, family, genus, species, AphiaID) %>%
  summarize(reads = sum(organismQuantity)) %>%
  ungroup()
```

## Resolve to accepted species

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
id_batches <- split(species$AphiaID, as.integer((seq_along(species$AphiaID) - 1) / 50))
resolved <- map(id_batches, wm_record) %>%
  bind_rows()

bad_ids <- resolved %>%
  filter(AphiaID != valid_AphiaID) %>%
  select(AphiaID, valid_AphiaID)

valid_taxa <- wm_record(bad_ids$valid_AphiaID) %>%
  filter(rank == "Species") %>%
  select(valid_AphiaID = AphiaID, phylum, class, order, family, genus, species = scientificname, scientificname) %>%
  inner_join(bad_ids, by = "valid_AphiaID")

species <- species %>%
  mutate(valid_AphiaID = NA_integer_) %>%
  rows_update(valid_taxa, by = "AphiaID") %>%
  mutate(AphiaID = ifelse(!is.na(valid_AphiaID), valid_AphiaID, AphiaID)) %>%
  select(-valid_AphiaID)
```

## Merge duplicate names

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
species <- species %>%
  group_by_at(vars(-reads)) %>%
  summarize(reads = sum(reads)) %>%
  ungroup()
```

## Add red list

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
redlist <- data.frame()
page <- 0
while (TRUE) {
  res <- rl_sp(page, key = "a936c4f78881e79a326e73c4f97f34a6e7d8f9f9e84342bff73c3ceda14992b9")$result
  if (length(res) == 0) {
    break
  }
  redlist <- bind_rows(redlist, res)
  page <- page + 1
}
redlist <- redlist %>%
  filter(is.na(population)) %>%
  filter(category %in% c("CR", "EN", "EW", "EX", "VU")) %>%
  select(species = scientific_name, category)

species <- species %>%
    left_join(redlist, by = "species")
```

## Compare with known species

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
mwhs_obis_species <- read.csv("species_lists/mwhs_species_list.csv") %>%
  rename(c("category" = "redlist_category")) %>%
  mutate(group = na_if(group, "")) %>%
  mutate(category = na_if(category, ""))

new_species <- species %>%
  filter(!AphiaID %in% mwhs_obis_species$AphiaID)

species <- species %>%
  mutate(new = ifelse(AphiaID %in% mwhs_obis_species$AphiaID, "", "yes"))
```

## Add vernacular names

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
vernaculars <- read.table("species_lists/vernacularname.txt", sep = "\t", quote = "", header = TRUE) %>%
  mutate(AphiaID = as.numeric(str_replace(taxonID, "urn:lsid:marinespecies.org:taxname:", ""))) %>%
  filter(language %in% c("FRA", "ENG")) %>%
  select(AphiaID, language, vernacularName) %>%
  group_by(AphiaID, language) %>%
  summarize(vernacularName = first(vernacularName)) %>%
  arrange(AphiaID, language) %>%
  group_by(AphiaID) %>%
  summarize(vernacular = paste0(vernacularName, collapse = ", "))

species <- species %>%
  left_join(vernaculars, by = "AphiaID")
```

## Add groups

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
fish_classes <- c("Actinopteri", "Cladistii", "Coelacanthi", "Elasmobranchii", "Holocephali", "Myxini", "Petromyzonti", "Teleostei")
turtle_orders <- c("Testudines")
mammal_classes <- c("Mammalia")

species <- species %>%
  mutate(
    group = case_when(
      class %in% fish_classes ~ "fish",
      order %in% turtle_orders ~ "turtle",
      class %in% mammal_classes ~ "mammal"
    )
  )
```

## Output species list

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
addtorow <- list()
addtorow$pos <- list()
addtorow$pos[[1]] <- c(0)
addtorow$command <- c(paste(
  "\\hline \n",
  "\\endhead \n",
  "\\hline \n",
  "{\\footnotesize Continued on next page} \n",
  "\\endfoot \n",
  "\\endlastfoot \n",
  sep = ""))

species_x <- xtable(species %>% select(phylum, class, species, group, category, new, vernacular), type = "latex")
print(species_x, file = "report/species.tex", include.rownames = FALSE, tabular.environment = "longtable", floating = FALSE, size = "\\fontsize{9pt}{10pt}\\selectfont", add.to.row = addtorow, hline.after = c(-1))
```

## Groups statistics

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
groups_obis <- mwhs_obis_species %>%
  filter(!is.na(group)) %>%
  group_by(group) %>%
  summarize(obis_species = n_distinct(species))

groups_edna <- species %>%
  filter(!is.na(group)) %>%
  group_by(group) %>%
  summarize(edna_species = n_distinct(species))

group_stats <- groups_obis %>%
  left_join(groups_edna, by = "group") %>%
  mutate(fraction = edna_species / obis_species)

group_stats_x <- xtable(group_stats, type = "latex")
caption(group_stats_x) <- "Number of species in the three target groups, from the OBIS database and from eDNA sampling."
label(group_stats_x) <- "table:groups"
print(group_stats_x, file = "report/group_stats.tex", include.rownames = FALSE, tabular.environment = "longtable", floating = FALSE)
```

## Threatened statistics

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
cat_obis <- mwhs_obis_species %>%
  filter(!is.na(category)) %>%
  filter(category %in% c("CR", "EN", "EW", "EX", "VU")) %>%
  group_by(category) %>%
  summarize(obis_species = n_distinct(species))

cat_edna <- species %>%
  filter(!is.na(category)) %>%
  group_by(category) %>%
  summarize(edna_species = n_distinct(species))

category_stats <- cat_obis %>%
  left_join(cat_edna, by = "category") %>%
  mutate(fraction = edna_species / obis_species)

category_stats_x <- xtable(category_stats, type = "latex")
caption(category_stats_x) <- "Number of species by IUCN Red List category, from the OBIS database and from eDNA sampling."
label(category_stats_x) <- "table:categories"
print(category_stats_x, file = "report/category_stats.tex", include.rownames = FALSE, tabular.environment = "longtable", floating = FALSE)
```

## Taxonomic tree with number of species and reads (top phyla)

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
library(data.tree)
library(tidyr)
library(ggtree)
library(ggtreeExtra) # BiocManager::install("ggtreeExtra")
library(phylobase)

df <- species %>%
  group_by(phylum, class, order, family) %>%
  summarize(species = n(), reads = sum(reads)) %>%
  mutate(domain = "Eukaryota") %>%
  ungroup()

top_phyla <- df %>%
  group_by(phylum) %>%
  summarize(species = sum(species)) %>%
  arrange(desc(species)) %>%
  head(10) %>%
  pull(phylum)

df_subset <- df %>%
  filter(phylum %in% top_phyla)

# marker stats

marker_stats <- occurrence %>%
  group_by(family, pcr_primer_name_forward) %>%
  summarize(reads = sum(organismQuantity)) %>%
  mutate(pcr_primer_name_forward = factor(pcr_primer_name_forward))

ggplot() + geom_point(data = marker_stats, aes(family, pcr_primer_name_forward, size = reads))

# dataframe to tree, look into alternatives

paths <- df_subset %>%
  unite(path, c(domain, phylum, class, order, family), sep = ";")
nwk <- ToNewick(as.Node(paths, pathName = "path", mode = "table", pathDelimiter = ";"))
writeLines(nwk, "tree.nwk")
tree <- read.tree("tree.nwk")

# add data to tree

# tree_data <- df %>%
#   select(species)
# row.names(tree_data) <- df$family
# tree_with_data <- phylo4d(tree, tip.data = tree_data)

# plot

ggtree(tree, layout = "circular") +
  geom_nodepoint() +
  geom_fruit(data = df_subset, aes(x = species, y = family, fill = phylum), geom = geom_bar, stat = "identity") +
  geom_fruit(data = marker_stats, aes(y = family, x = pcr_primer_name_forward, size = reads, color = pcr_primer_name_forward), geom = geom_point, name = "marker") +
  geom_tiplab(size = 3, offset = 60) +
  scale_fill_brewer(palette = "Spectral") +
  scale_color_brewer(palette = "Paired") +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend(title = "marker"))

ggsave("report/tree.png", scale = 1.2, width=10, height=9)
```
